Strict Pure Functional Programming and Automatic Parallelization and Transparent Fault Tolerance\\ (New-project-in-progress Paper)

### 2016-03-25 pubs/tfp16/paper.tex
mac->dos->unix
minor edits

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
% mkd for preliminary paper notes
\usepackage{enumitem} 
\usepackage{color} % for parenthetical notes while paper is in progress
%
\begin{document}
%
%mkd \frontmatter          % for the preliminaries
%
\pagestyle{headings}  % switches on printing of running heads
%
%mkd \mainmatter              % start of the contributions
%
\title{Strict Pure Functional Programming and Automatic Parallelization\\
       (New-project-in-progress Paper)}
%
\titlerunning{Automatic Parallelization}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Kei Davis\inst{1} \and
        Dean Prichard\inst{1} \and
        David Ringo\inst{1,2} \and
        Loren Anderson\inst{1,3} \and
        Jacob Marks\inst{1,4}\thanks{All project contributors, past and present, are listed.}
}
%
\authorrunning{Kei Davis et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
%
\institute{Los Alamos National Laboratory, Los Alamos, NM, USA\\
\email{kei.davis@lanl.gov},\\ WWW home page:
\texttt{http://ccsweb.lanl.gov/\homedir kei/}
\and
University of New Mexico,
Albuquerque, NM, USA
\and
North Dakota State University,
Fargo, North Dakota, USA
\and
New Mexico Institute of Mining and Technology,
Socorro, New Mexico, USA
}

@@

\begin{abstract}
The claimed trend is an increasing appreciation and practice of functional
programming in scientific computing, and in particular strict (by default)
functional programming, in various guises.  Our project is the demonstration
of a light-weight, higher-order, polymorphic, pure functional language
implementation in which we can experiment with automatic parallelization
strategies and varying degrees of default strictness.  Non-strictness
constrains the degree of potential parallelism, that is, we are not exploring
speculative evaluation.
\\\ \\
\emph{Preliminatry paper notes:}
\begin{enumerate}[before=\itshape]
  \item We are reporting work in progress.  At the time of writing we have a
    complete serial implementation, a primitive proof-of-concept parallel
    implementation, and are currently developing a more realistic parallel
    runtime.  We expect to have preliminary results from our new parallel
    implementation by the time of the workshop.

  \item While this does not qualify as a student paper, students
    (undergraduate!)  have contributed considerably to this effort, and if
    accepted we plan to have one of these students present this paper.

@@

\section{Background and Motivation}
The practice of so-called high-performance scientific computing (HPC) is,
overall, highly conservative with respect to change.  Even as the US
government commits to a push for exascale computing ($10^{18}$ FLOPS in a
single system) within the next decade, a highly respected member of the HPC
community recently stated publicly that ``Fortran is essential for exascale
programming.''~\cite{Heroux16}\footnote{Various EU, Japanese, and
  international groups are also anticipating exascale
  computing.~\cite{EUexascale,Exascale-org}} Existing US government laboratory
HPC codes run to hundreds of thousands of lines of Fortran and complete
rewrites are infeasible.~\footnote{The codes of our lab's British counterpart
  are reportedly in the same state.}  C++ has become more common for new code
starts, but their nature remains the same: huge codes with fair to
non-existent encapsulation of side effects, with multiple levels of
parallelism also poorly abstracted, especially at the thread (shared-memory)
level.  Thread-level parallel efficiency can be poor and is explained by
Amdahl's law because the parallel model \emph{as typically implemented} is
bulk synchronous.

@@

Much anecdotal evidence suggests that many others have similar difficulties
with lazy (or non-strict) by default semantics, with the most telling recent
concrete evidence being the push for, and realization of, the \texttt{Strict}
and \texttt{StrictData} language extensions newly available in GHC 8.0.
Quoting the documentation:
\begin{quote}
High-performance Haskell code (e.g. numeric code) can sometimes be littered
with bang patterns, making it harder to read. The reason is that laziness
isn't the right default in this particular code, but the programmer has no way
to say that except by repeatedly adding bang patterns~\cite{strict-strictdata}.
\end{quote}

@@

Intel's TBB?  Cilk++?

@@

Our project is the light-weight implementation of a pure, higher-order,
polymorphic, functional language and runtime system with which we can
experiment with automatic parallelization strategies with degrees of language
strictness ranging from lazy to fully strict.  Light-weight has several
meanings or implications.  First and foremost it means that \emph{we are not
  attempting to compete with ghc in any respect}.\footnote{We estimate that we
  will have, to within an order of magnitude, one person-hour of available
  development time for each person-year that ghc currently represents, or put
  another way, one second for each hour, respectively.}

\begin{itemize}
\item The implementation must be feasible in terms of effort;
\item The language must be strongly typed, implicitly or explicitly.  More
  specifically, the language should explicitly support both Hindley-Milner
  polymorphism and unboxed types~\cite{Jones:1991};
\item Straightforward direct linkage with C, preferably without a sophisticated
  foreign function interface, is essential;
  \item The language should be \emph{currying-friendly} as described by Marlow
and Peyton Jones~\cite{Marlow2004};
\item Choice of degree of strictness is easily selected;
\item Sharing, in the sense of what \emph{lazy} means vs.\ merely non-strict,
  should be preserved for all degrees of strictness;
\item Tail calling should be properly implemented, i.e., not as a trampoline;
\item etc.
\end{itemize}

@@

A secondary, informal goal is to gain a sense of how often relatively
inexperienced functional programmers make essential use of
non-strictness.\footnote{The starting experience of our student interns to
  date has been normalized by having them complete Hutton's \emph{Programming in Haskell} textbook
as baseline preparation.~\cite{Hutton-book}}

@@

Our wish list sounds much like the ghc Haskell implementation, and indeed ghc
is part of our master plan.  In short, our plan is to use ghc as our front-end
to generate its intermediate form, STG, then escape to our system from there.
There are various possible ways to accomplish the former: using ghc's option
to dump STG as text (see Stuart's report~\cite{Stuart}), hacking ghc itself
and, optionally, using ghc as a library.  We note that this approach has
a precedent in the aforementioned Nvidia research compiler.

@@

Other possible variations concern the notional arity of the underlying
function being applied.  This is readily explained in terms of the operation
of the STG machine described later.  Our description is based on that given in
Marlow and Peyton Jones' \emph{fast curry} paper~\cite{fastcurry}, to which we
refer so many times that we will subsequenty refer to it as
\emph{EA}.\footnote{\emph{Eval/Apply} from the title because \emph{FC} likely
  has another connotation to the reader.}

@@

The back-end is written in Haskell and generates proper tail-calling C code;
the runtime system is written in C/C++.  The system fairly faithfully
implements the STG machine as described in Peyton-Jones and Marlow's ``fast
curry'' paper~\cite{fastcurry} and dynamically illustrated by Pope's
miniSTG~\cite{ministg} with the exception of the state transition rules which,
of course, are different than those for lazy semantics.  Sharing is preserved.

@@

Proper tail calling is achieved.  Because we maintain our own data and control
(continuation) stack all generated C functions take no arguments and have
return type void, and with a single exception in the non-strict case are
invoked in tail call positions, at non-zerop optimization levels both gcc and
Clang/LLVM generate jumps instead of calls for C calls in tail call positions.

The exceptions to tail calling are for convenience.  Similar to what is
described in the \emph{fast curry} paper we have a generic variadic stgApply
function that is used for applications of non-known functions, and known
functions with a deficit or excess number of arguments.  The stgApply function
takes as arguments the first element of the application (FUN, PAP, THUNK, or
BLACKHOLE) and its arguments, and if a THUNK, evaluates it with a call (rather
than jump) to yield a FUN or PAP.  This allows us to implement a single
stgApply function as a single C function.  The implementation of stgApply as
described in \emph{fast curry} makes the same concession.

For stgApply in the non-strict semantics the alternative to call-with-return
would be a pair of functions stgApply1 and stgApply2.  In one possible variation,
stgApply1 would push a continuation with stgApply2 as the return address, then
invoke the first element of the application.  Call-with-return becomes more
convenient when considering the various forms of strictness:  when evaluating
function or constructor arguments it is simpler, and arguably more efficient
in terms of time (but not space in general) to ``evaluate in place'' than
implement the continuation style to the logical limit.

@@

We have implemented a simple copying garbage collector \cite{Cheney:1970}
with small "nurseries" in the multithreaded case \cite{Marlow:2008}. We have not
focused on the development of a more sophisticated garbage collector, as the goal
of this project is to explore levels of strictness.

@@

We perform standard Hindley-Milner type inference and enforcement, augmented
with built-in and user-defined unboxed types as described by Peyton-Jones and
Launchbury's seminal work~\cite{unboxed}, but for unboxed types code
generation is currently limited to built-in single-word (64-bit) or smaller
types (e.g., \texttt{Int\#}, \texttt{UInt\#}, \texttt{Float\#}, \texttt{Double\#})
and user-defined enumerations, e.g., \texttt{data unboxed Bool\# = False\# |
  True\#}.

@@

\section{Other Related Work}
Nvidia research compiler.

@@

Funding for this project and its participants has been provided by the DOE
NNSA LANL Laboratory Directed Research and Development program award
20150845ER, the National Science Foundation Science, Technology, Engineering,
and Mathematics Talent Expansion Program (STEP) program for undergraduate
students, and the DOE Science Undergraduate Laboratory Internship program for
undergraduate students.
%
Los Alamos National Laboratory is managed and operated by Los Alamos National
Security, LLC (LANS), under contract number DE-AC52-06NA25396 for the
Department of Energy’s National Nuclear Security Administration (NNSA).

>>>>

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
% mkd for preliminary paper notes
\usepackage{enumitem} 
\usepackage{color} % for parenthetical notes while paper is in progress
%
\begin{document}
%
%mkd \frontmatter          % for the preliminaries
%
\pagestyle{headings}  % switches on printing of running heads
%
%mkd \mainmatter              % start of the contributions
%
\title{Strict Pure Functional Programming and Automatic Parallelization and Transparent Fault Tolerance\\
       (New-project-in-progress Paper)}
%
\titlerunning{Automatic Parallelization}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Kei Davis\inst{1} \and
        Dean Prichard\inst{1} \and
        David Ringo\inst{1,2} \and
        Loren Anderson\inst{1,3} \and
        Jacob Marks\inst{1,4}\thanks{All project contributors, past and present, are listed.}
}
%
\authorrunning{Kei Davis et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
%
\institute{Los Alamos National Laboratory, Los Alamos, NM, USA\\
\email{kei.davis@lanl.gov},\\ WWW home page:
\texttt{http://ccsweb.lanl.gov/\homedir kei/}
\and
University of New Mexico,
Albuquerque, NM, USA
\and
North Dakota State University,
Fargo, North Dakota, USA
\and
New Mexico Institute of Mining and Technology,
Socorro, New Mexico, USA
}

@@

\begin{abstract}
The claimed trend is an increasing appreciation and practice of functional
programming in scientific computing, and in particular strict (by default)
functional programming, in various guises.  Our project is the demonstration
of a light-weight, higher-order, polymorphic, pure functional language
implementation in which we can experiment with automatic parallelization
strategies and varying degrees of default strictness.  Non-strictness
constrains the degree of potential parallelism, that is, we are not exploring
speculative evaluation.  A secondary goal is to experiment with mechanisms
for transparent fault tolerance.
\\
\emph{Preliminatry paper notes:}
\begin{enumerate}[before=\itshape]
  \item We are reporting work in progress.  At the time of writing we have a
    complete serial implementation, a primitive proof-of-concept parallel
    implementation, and are currently developing a more realistic parallel
    runtime.  We expect to have preliminary results from our new parallel
    implementation by the time of the workshop.  The fault-tolerance mechanisms
    remain future work.

  \item While this does not qualify as a student paper, students
    (undergraduate!)  have contributed considerably to this effort, and to the
    writing of this paper, and will continue to do so.  If this paper is
    accepted we plan to have one of these students present it at TiFP.

@@

\section{Background and Motivation}
The practice of so-called high-performance scientific computing (HPC) is,
overall, highly conservative with respect to change.  Even as the US
government commits to a push for exascale computing ($10^{18}$ FLOPS in a
single system) within the next decade,~\cite{Exascale-roadmap} a highly
respected member of the HPC community recently stated publicly that ``Fortran
is essential for exascale programming.''~\cite{Heroux16}\footnote{Various EU,
  Japanese, and international groups are also anticipating the realization
  exascale computing.~\cite{EUexascale,Exascale-org}} Existing US government
laboratory HPC codes run to hundreds of thousands of lines of Fortran and
complete rewrites are infeasible.~\footnote{The codes of our lab's British
  counterpart are reportedly in the same state.}  C++ has become more common
for new code starts, but their nature remains the same: huge codes with fair
to non-existent encapsulation of side effects, with multiple levels of
parallelism also poorly abstracted, especially at the thread (shared-memory)
level.  Thread-level parallel efficiency can be poor and is explained by
Amdahl's law because the parallel model \emph{as typically implemented} is
bulk synchronous.

@@

Much anecdotal evidence suggests that many others have similar difficulties
with lazy (or non-strict) by default semantics, with the most telling recent
concrete evidence being the push for, and realization of, the \texttt{Strict}
and \texttt{StrictData} language extensions newly available in GHC 8.0.
Quoting the documentation:
\begin{quote}
  High-performance Haskell code (e.g.\ numeric code) can sometimes be littered
  with bang patterns, making it harder to read. The reason is that laziness
  isn't the right default in this particular code, but the programmer has no
  way to say that except by repeatedly adding bang
  patterns~\cite{strict-strictdata}.
\end{quote}
It is a fact that \emph{HPC practioners are by and large not computer scientists}.
The concept of a space leak, for example, caused by other than a missing explicit
deallocation statement (C \emph{free}, etc.) is difficult to understand and not
something they particularly want to think about.  Hence, we eschew explicit
strictness annotations for varying uniform degrees of strictness.

@@

More generally, \emph{task-based} parallelism is becoming increasingly popular
for HPC, often in conjunction with a higher-level (typically inter-node) bulk
synchronous model typically implemented with the Message Passing Interface
(MPI)~\cite{MPI}.  In the imperative world, example systems include the
aforementioned Legion, Intel's Threading Building Blocks~\cite{Reinders:2007}, and the
HPX C++11/14 extension~\cite{HPX}.  While the concept of task is somewhat
vague, the general idea is a unit of computation with no or
minimal/constrained side effects.  Function call in a pure functional language
could be regarded as the essence of the concept of task.

In the quest for computing platforms approaching exascale, it is widely
recognized that new mechanisms for fault tolerance will need to be built into
the software stack---the scalability of checkpoint/restart may be reaching its
end.  In the imperative world side-effect-free tasks could be safely restarted
after failure, and the Legion system is being augmented with this capability.
Ericsson's quasi-functional language Erlang and its runtime system was
designed for fault tolerance.~\cite{Cesarini:2009}.  It has long been recognized
that the pure parts of a functional program (excluding, e.g., the I/O monad
in Haskell) could be safely restarted.

@@

Our project is the light-weight implementation of a pure, higher-order,
polymorphic, functional language and runtime system with which we can
experiment with automatic parallelization strategies with degrees of language
strictness ranging from lazy to fully strict.  Light-weight has several
meanings or implications.  First and foremost it means that \emph{we are not
  attempting to compete with ghc in any respect}---succumbing to the
temptation of recreating ghc language, type system, and runtime features, or
significant code optimization, would simply divert us from our central
goals.\footnote{We estimate that we will have, to within an order of
  magnitude, one person-hour of total available development time for each
  person-year that ghc currently represents, or put another way, one second
  for each hour, respectively.}

\begin{itemize}
\item The implementation must be feasible in terms of effort;
\item The language must be strongly typed, implicitly or explicitly.  More
  specifically, the language should explicitly support both Hindley-Milner
  polymorphism and unboxed types~\cite{Jones:1991};
\item Straightforward direct linkage with C, preferably without a sophisticated
  foreign function interface, is essential;
  \item The language should be \emph{currying-friendly} as described by Marlow
and Peyton Jones~\cite{Marlow:2004};
\item Choice of degree of strictness is easily selected;
\item Sharing, in the sense of what \emph{lazy} means vs.\ merely non-strict,
  should be preserved for all degrees of strictness;
\item Tail calling should be properly implemented, i.e., not as a trampoline;
\item etc.
\end{itemize}

@@

A secondary, informal goal is to gain a sense of how often relatively
inexperienced functional programmers make essential use of
non-strictness.\footnote{The starting experience of our student interns to
  date has been normalized by having them complete Hutton's \emph{Programming
    in Haskell} textbook~\cite{Hutton-book} as baseline preparation,
  substituting his article \emph{Higher-order Functions for
    Parsing}~\cite{Hutton-parsing:1992} for the corresponding material in the
  book.}

@@

Our wish list sounds much like the ghc Haskell implementation, and indeed ghc
is part of our master plan.  In short, our plan is to use ghc as our front-end
to generate its intermediate form, STG, then escape to our system from there.
There are various possible ways to accomplish the former: using ghc's option
to dump STG as text, hacking ghc itself
and, optionally, using ghc as a library.  We note that this approach has
a precedent in the aforementioned Nvidia research compiler.

@@

Other possible variations concern the notional arity of the underlying
function being applied.  This is readily explained in terms of the operation
of the STG machine described later.  Our description is based on that given in
Marlow and Peyton Jones' \emph{fast curry} paper~\cite{Marlow:2004}, to which we
refer so many times that we will subsequenty refer to it as
\emph{EA}.\footnote{\emph{Eval/Apply} from the title because \emph{FC} likely
  has another connotation to the reader.}

@@

The back-end is written in Haskell and generates proper tail-calling C code;
the runtime system is written in C/C++.  The system fairly faithfully
implements the STG machine as described in \emph{EA} and dynamically
illustrated by Pope's Ministg~\cite{ministg} with the exception of the state
transition rules which, of course, may be different than those for non-strict
semantics.  Sharing is preserved.

@@

While performance is not a high priority, we attain proper tail calling instead of
using the trampoline used in earlier versions of ghc~\cite{PJ-stockhardware}.
Because we maintain our own data and control (continuation) stack all
generated C functions take no explicit arguments and have return type void,
and with a single exception in the non-strict case are invoked in tail call
positions, at non-zero optimization levels both gcc and Clang/LLVM generate
jumps instead of calls for C calls in tail call positions.  When using gcc
we also use a fixed set of machine registers for passing/returning values.

The exceptions to tail calling are for convenience.  Similar to what is
described in the \emph{EA} we have a generic variadic stgApply function that
is used for applications of non-known functions, and known functions with a
deficit or excess number of arguments.  The stgApply function takes as
arguments the first element of the application (FUN, PAP, THUNK, or BLACKHOLE)
and its arguments, and if a THUNK, evaluates it with a call (rather than jump)
to yield a FUN or PAP.  This allows us to implement a single stgApply function
as a single C function taking a contiguous block of machine words (unboxed
values or pointers) on the explicitly stack.  The implementation of stgApply
as described in \emph{EA} makes the same concession regarding
call-with-return, but the scheme described there uses an auto-generated family
of stgApply functions parameterized by number and type (unboxed or pointer) of
arguments.  We used this scheme initially as well, but because the parameter
block contains a bitmap describing which arguments are boxed/unboxed for the
benefit of the garbage collector, this elaboration is unnecessary.

For stgApply in the non-strict semantics an alternative to call-with-return
would be a pair of functions stgApply1 and stgApply2.  In one possible variation,
stgApply1 would push a continuation with stgApply2 as the return address, then
invoke the first element of the application.  Call-with-return becomes more
convenient when considering the various forms of strictness:  when evaluating
function or constructor arguments it is simpler, and arguably more efficient
in terms of time (but not space in general) to ``evaluate in place'' than
implement the continuation style to the logical limit.

@@

We have implemented a simple copying garbage collector \cite{Cheney:1970} with
small \emph{nurseries} in the multithreaded case \cite{Marlow:2008}. We have
not focused on the development of a more sophisticated garbage collector
because the goal of this project is to explore automatic parallelization.

@@

We perform standard Hindley-Milner type inference and enforcement, augmented
with built-in and user-defined unboxed types as described by Peyton-Jones and
Launchbury's seminal work~\cite{Jones91unboxedvalues}, but for unboxed types code
generation is currently limited to built-in single-word (64-bit) or smaller
types (e.g., \texttt{Int\#}, \texttt{UInt\#}, \texttt{Float\#}, \texttt{Double\#})
and user-defined enumerations, e.g., \texttt{data unboxed Bool\# = False\# |
  True\#}.

@@

\section{Other Related Work}

In his BSc thesis, Don Stewart described an implementation wherein ghc was used as a Haskell
compiler front-end, dumping STG code as text, parsing that text, then translating to code for
execution on the Java Virtual Machine.~\cite{Stewart-BSc}.

Nvidia research compiler.

@@

Funding for this project and its participants has been provided by the DOE
NNSA LANL Laboratory Directed Research and Development program award
20150845ER; the National Science Foundation Science, Technology, Engineering,
and Mathematics Talent Expansion Program (STEP) program for undergraduate
students; and the DOE Science Undergraduate Laboratory Internship (SULI) program.
%
Los Alamos National Laboratory is managed and operated by Los Alamos National
Security, LLC (LANS), under contract number DE-AC52-06NA25396 for the
Department of Energy’s National Nuclear Security Administration (NNSA).
